---
title: "Edge - Array analysis of inner ear cells"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 4
    fig_caption: true
    fig_width: 8
    fig_height: 6
author: "Meeta Mistry"
---

```{r setup, echo=FALSE}

# Setup report details
clientname="Judith Kempfle"
clientemail="Judith_Kempfle@MEEI.HARVARD.EDU"
lablocation="HSCI - Edge Lab"
analystname="Meeta Mistry"
analystemail="mmistry@hsph.harvard.edu"
```

Array analysis for `r clientname` (`r clientemail`) at `r lablocation`. Contact `r analystname` (`r analystemail`) for additional details. Request from client was:

> Arrays run on Mouse Ref 8 v2.0 Expression Array Ambion


## Workflow: 
* 
* 
* 
* 


## Setup

### Bioconductor and R libraries used

```{r libraries, echo=TRUE}
library(IDATreader) ## must have base64 and Rtools installed
library(Biobase)
library(limma)
library(lattice)
library(ggplot2)
library(RColorBrewer)
library(arrayQualityMetrics)
library(reshape)
library(dplyr)
library(CHBUtils)
```

### Get variables
- get base directory for analyses
- specify data and results directories
- specify column headers used in metadata file


```{r variables, echo=TRUE}
# Setup directory variables
baseDir <- '.'
dataDir <- file.path(baseDir, "data")
metaDir <- file.path(baseDir, "meta")
resultsDir <- file.path(baseDir, "results")
```


## Load data: Decrypt IDAT files into an expression matrix
```{r idatreader}

# Get path for all .idat files
dirlist <- scan(file.path(dataDir, "directories.txt"))
filenames = lapply(dirlist, function(x){
    path <- file.path(dataDir, x)
    files <- list.files(path=path, pattern=".idat")
    return(paste0(path, "/", files))})
filenames <- unlist(filenames)

# Decrypt the .idat files
theData = lapply(filenames, readIDAT)

# Extract expression information only
exprs = sapply(theData, function(x) {
  out <- x$TrimmedMeanBinData
  names(out) <- x$CodesBinData
  return(out)})

# Get column names for expression matrix
getcolnames = lapply(filenames, function(x){
  sp <- strsplit(x, "/", fixed=T)
  s <- unlist(strsplit(sp[[1]][4], "_", fixed=T))
  cname <- paste(s[1], s[2], sep="_")
  return(cname)})
colnames(exprs)= unlist(getcolnames)

# Map rownames to Illumina IDs
ref8 <- read.delim(file.path(metaDir, "MouseRef-8_V2.txt"), sep="\t", header=T,)
ref8ctl <- read.csv(file.path(metaDir, "control probes.csv"), header=T)

annot <- ref8[which(ref8$Array_Address_Id %in% rownames(exprs)), c('Array_Address_Id', 'Probe_Id', 'ILMN_Gene',
                                                                   'Entrez_Gene_ID', 'Cytoband')]
annot_ctl <- ref8ctl[which(ref8ctl$ProbeID %in% rownames(exprs)), c('ProbeID', 'TargetID')]

# Missing 292 probe annotations
pids<-c(annot$Array_Address_Id, annot_ctl$ProbeID)
exprs.new <- exprs[which(rownames(exprs) %in% pids),]
```

## Background correction (using neqc::limma)
Illumina negative control probes (sequences with no match to the genome) can be used to correct observed signal intensity using a normal exponential convolution model to reduce bias and FPS. Adding small offset to corrected intensities can reduce FDR further (neqc offset = 16).

```{r bg-correct}

neg <- annot_ctl$ProbeID
reg <- annot$Array_Address_Id
status <- rownames(exprs.new)
status <- replace(status, which(status %in% neg == TRUE), "negative")
status <- replace(status, which(status %in% reg == TRUE), "regular")

#estimating proportion of detected probes
exp.prop <- propexpr(exprs.new, status=status, labels=c("negative", "regular") )
prop.low <- names(exp.prop)[which(exp.prop <= 0.2)]

#parameters required by normal+exponential(normexp) convolution model with the help of negative control probes
param <- normexp.fit.control(exprs.new, status=status, negctrl="negative", regular="regular", robust=FALSE)

#background correction followed by quantile normalization and log2 transformation
exprs.bg <- neqc(exprs.new, status=status, negctrl="negative", regular="regular", offset=16)# returns matrix without control probes
```


## Create expression set object
```{r eset}

# eset <- new("ExpressionSet", phenoData=as(pheno, "AnnotatedDataFrame"), exprs=as.matrix(exprs.bg))
eset <- new("ExpressionSet", exprs=as.matrix(exprs.bg))
fData(eset) <- data.frame(annot[,2:5], row.names=annot$Array_Address_Id)

```





