---
title: "Edge - Array analysis of inner ear cells"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 4
    fig_caption: true
    fig_width: 8
    fig_height: 6
author: "Meeta Mistry"
---

```{r setup, echo=FALSE}

# Setup report details
clientname="Judith Kempfle"
clientemail="Judith_Kempfle@MEEI.HARVARD.EDU"
lablocation="HSCI - Edge Lab"
analystname="Meeta Mistry"
analystemail="mmistry@hsph.harvard.edu"
```

Array analysis for `r clientname` (`r clientemail`) at `r lablocation`. Contact `r analystname` (`r analystemail`) for additional details. Request from client was:

> Arrays run on Mouse Ref 8 v2.0 Expression Array Ambion


## Workflow: 
* 
* 
* 
* 


## Setup

### Bioconductor and R libraries used

```{r libraries, echo=TRUE}
library(IDATreader) ## must have base64 and Rtools installed
library(Biobase)
library(limma)
library(lattice)
library(ggplot2)
library(RColorBrewer)
library(arrayQualityMetrics)
library(reshape)
library(dplyr)
library(CHBUtils)
```

### Get variables
- get base directory for analyses
- specify data and results directories
- specify column headers used in metadata file


```{r variables, echo=TRUE}
# Setup directory variables
baseDir <- '.'
dataDir <- file.path(baseDir, "data")
metaDir <- file.path(baseDir, "meta")
resultsDir <- file.path(baseDir, "results")
```


## Load data: Decrypt IDAT files into an expression matrix
```{r idatreader}

# Get path for all .idat files
dirlist <- scan(file.path(dataDir, "directories.txt"))
filenames = lapply(dirlist, function(x){
    path <- file.path(dataDir, x)
    files <- list.files(path=path, pattern=".idat")
    return(paste0(path, "/", files))})
filenames <- unlist(filenames)

# Decrypt the .idat files
theData = lapply(filenames, readIDAT)

# Extract expression information only
exprs = sapply(theData, function(x) {
  out <- x$TrimmedMeanBinData
  names(out) <- x$CodesBinData
  return(out)})

# Get column names for expression matrix
getcolnames = lapply(filenames, function(x){
  sp <- strsplit(x, "/", fixed=T)
  s <- unlist(strsplit(sp[[1]][4], "_", fixed=T))
  cname <- paste(s[1], s[2], sep="_")
  return(cname)})
colnames(exprs)= unlist(getcolnames)

# Map rownames to Illumina IDs
ref8 <- read.delim(file.path(metaDir, "MouseRef-8_V2.txt"), sep="\t", header=T,)
ref8ctl <- read.csv(file.path(metaDir, "control probes.csv"), header=T)

m <- match(row.names(exprs), ref8$Array_Address_Id)

annot <- ref8[which(ref8$Array_Address_Id %in% rownames(exprs)), c('Array_Address_Id', 'Probe_Id', 'ILMN_Gene')]
annot_ctl <- ref8ctl[which(ref8ctl$ProbeID %in% rownames(exprs)), c('ProbeID', 'TargetID')]
annot_ctl <- cbind(annot_ctl, rep(NA, nrow(annot_ctl)))
colnames(annot_ctl) <- colnames(annot)

sub<-HT12ctl[which(HT12ctl$Array_Address_Id %in% rownames(exprs) == T),]
annot<-rbind(annot, cbind(sub$Array_Address_Id, sub$Probe_Id))
rownames(annot)<-annot[,1]
annot<-annot[,-1]

pids<-annot[row.names(exprs)]
exprs<-exprs[-which(is.na(pids)),]
rownames(exprs)<-pids[which(!is.na(pids))]
```


